---
title: "PATH analysis of somatic evolution"
output: 
  html_document:
    fig_height: 3
    fig_width: 7
    fig_retina: 2
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center"
)
```

```{r setup, message=FALSE}
library(PATH)
library(expm)
library(ggplot2)
library(ggtree)
library(ggtreeExtra)
library(tidytree)
library(Matrix)
library(patchwork)

options(digits = 2)
```

### Introduction

In this vignette we demonstrate how to use PATH to measure categorical cell state 
phylogenetic correlations and how to infer cell state transition dynamics
on phylogenies simulated as a *sampled somatic evolutionary process*. 
For a similar analysis using *idealized* phylogenies, or a more in depth
introduction of how we model cell state transition dynamics, see 
[this vignette](./Idealized_phylogenies.html). 
For more information, see our [bioRxiv preprint](). 

### Somatic evolution {#somatic}

Here, we demonstrate PATH as in the [section](#ideal) above, but instead of
analyzing idealized phylogenies, we use phylogenies simulated as a sampled
somatic evolutionary process. The process used to simulate these phylogenies,
while more complex than the one above, is likely a better representation of
somatic evolution, as cell birth and death occur stochastically, and the
phylogeny represents a sample of the total somatic population. 

Here, we first randomly generate
the transition rate matrix $Q$ using the PATH function `rQ()`,
by specifying the number of cell states $n$ with the function's first argument. 
Here we use this function to randomly generate a $3$-dimensional 
transition rate matrix and use `expm` to get the transition probability matrix. 

#### Simulate
We then use `ran_tree_with_states()` to simulate a phylogeny using a sampled
somatic evolutionary process, and set the number of sampled cells to be
`N = 1000`, with a sampling rate of `rho = 10^-3`, meaning that the total
simulated somatic population contains one million extant cells at the time of
sampling. 

```{r }

Q <- rQ(3, 0.1)
P <- expm(Q)
# Simulate phylogeny (phy1) with a sampled somatic evolutionary process
# using specified transition rate matrix (Q0), cell number (N), and
# cell sampling fraction (rho).
phy1 <- ran_tree_with_states(Q0 = Q, N = 1000, rho = 10^-3)
```

We can visualize this phylogeny and annotate the cells by their state. 

```{r, fig.height=5, echo=FALSE}
# Format simulated phylogeny to use ggtree for visualization
# by converting phy1 to phy1_tidy.
phy1_tidy <- full_join(phy1, 
                          tibble(label=phy1$tip.label, 
                                 "state"=as.factor(phy1$states)),
                          by="label") 

# Use ggtree to plot phylogeny and annotate leaves (terminal nodes) by 
# categorical cell state.
ggtree(phy1_tidy, layout = "circular", size=0.1) + 
  geom_fruit(geom = "geom_tile", aes(fill=state), 
             col="white", offset = 0.15, 
             width = 3, lwd = 0) +
  scale_fill_brewer(palette = "Set1", type = "div") +
  ggtitle("phy1 phylogeny") + labs(fill="Cell state")
```

#### Measure phylogenetic correlations
As was done [here](./Idealized_phylogenies.html), 
we compute phylogenetic correlations
using `xcor()`, after
first computing a normalized phylogenetic weight matrix with
`one_node.tree.dist()`, and by reformatting the cell state vector
into a matrix with `state2mat.sparse()`.

```{r }
# Get phylogenetic node distance matrix 
# retaining only node distances of 1.
W <- one_node.tree.dist(phy1)

# Transform vector of cell state assignments into matrix
# format for calculating phylogenetic correlations. 
X <- state2mat.sparse(phy1$states)

# Compute phylogenetic correlations with xcor().
# Output of xcor() is a list,  including:
# phylogenetic correlations ("Morans.I"), and
# leaf-permutation based
# analytical z scores ("Z.score"), 
# variance ("Var.I"), expected values ("Expected.I"),
# and one-sided p-values ("one.sided.pvalue").
phy1xcor <- xcor(X, W)
```

```{r, echo=FALSE}
# Process xcor output for plotting and visualization.
Idf <- reshape2::melt(phy1xcor$Morans.I, 
                      value.name = "I")
Zdf <- reshape2::melt(phy1xcor$Z.score, 
                      value.name = "Z")

df <- full_join(Idf, Zdf, by=c("Var1", "Var2"))

df <- df %>% mutate(Var1 = as.factor(Var1), 
                      Var2 = as.factor(Var2))
```

We use phylogenetic auto-correlations as a measure of cell state clustering, 
or heritability, and the concordant z scores to assess statistical significance. 
Both auto-correlations ``diag(phy0xcor$Morans.I)`` and 
z scores ``diag(phy0xcor$Z.scores)`` are plotted below. 

```{r, fig.width=6, echo=FALSE}
# Phylogenetic auto-correlation bar plot.
# df is a reformatted data.frame of xcor() output from above.
herit1 <- df %>%
  filter(Var1 == Var2) %>%
  ggplot(aes(x=Var1, y=I, fill=Var1)) +
  geom_bar(stat="identity") +
  ylim(c(-1,1)) +
  theme_classic() +
  scale_fill_brewer(palette = "Set1", type = "div") +
  labs(fill="Cell state") +
  ylab("Phylogenetic auto-correlation") + 
  xlab("Cell state") +
  ggtitle("Cell state heritability", 
          subtitle = "auto-correlation")

maxz <- max(abs(df$Z))
herit2 <- df %>%
  filter(Var1 == Var2) %>%
  ggplot(aes(x=Var1, y=Z, fill=Var1)) +
  geom_bar(stat="identity") +
  ylim(c(-maxz,maxz)) +
  theme_classic() +
  scale_fill_brewer(palette = "Set1", type = "div") +
  labs(fill="Cell state") +
  ylab("Phylogenetic auto-correlation\n(z score)") + 
  xlab("Cell state") +
  geom_hline(yintercept = qnorm(0.05, lower.tail = F), 
             col="black", lty=2) +
  ggtitle("Cell state heritability", 
          subtitle = "z score")

herit1 + herit2 + plot_layout(guides = "collect")  

```

The phylogenetic relationships between cell states as quantified with
phylogenetic cross-correlations ``phy0xcor$Morans.I`` can be visualized 
as a heat map, as below. Note that the diagonal of the heat map represents 
the same values as depicted in the auto-correlation bar plot above. 

```{r, echo=FALSE}
# Plot phylogenetic correlation heat map.
df %>%
  ggplot(aes(x=Var1, y=Var2, fill=I)) +
  geom_tile(col="white") +
  scale_fill_distiller(palette = 5, type = "div",
                       limits=c(-1,1)) +
  theme_classic() +
  scale_y_discrete(limits=rev) +
  labs(fill="Phylogenetic\ncorrelation") +
  xlab("Cell state") + ylab("Cell state") +
  theme(aspect.ratio = 1)

```

#### Infer cell state dynamics
Finally, we can use these phylogenetic correlations to infer cell state
state transition dynamics with the function `PATH.inference()`. This function
takes a phylogeny (here, `phy0`), the name of the cell state vector 
(here, `"states"` as cell states are stored at `phy0$states`), and `nstates`,
the number of possible cell states.

We use two different inference approaches. The first inference uses *measured*, 
and the second uses *imputed* (`imputed_branches = TRUE`),
branch length distances (the distance in units of time between cells
separated by a node distance of one). Branch length imputation can be useful
if we suspect that measured branch lengths are inaccurate or if missing. 

```{r }
# Infer cell state transition probabilities from phylogenetic
# correlations and measured branch lengths using PATH.
Pinf <- PATH.inference(tree = phy1, 
                              cell_states = "states", 
                              nstates = nrow(P))

# Infer cell state transition probabilities from phylogenetic
# correlations and imputed branch lengths using PATH.
# Note that our sampling rate estimate (sample_rate_est) of 10^-2 
# is incorrect, as the simulated (true) sampling rate is 10^-3. 
Pinf.impute <- PATH.inference(tree = phy1, 
                                     cell_states = "states", 
                                     nstates = nrow(P), 
                                     impute_branches = TRUE, 
                                     sample_rate_est = 10^-2)

```


First we can see what the inferred transition probabilities are for a length
of time proportional to the branch length distances separating cell pairs one
node apart. 
```{r}
# PATH-inferred transitions for time corresponding to a node depth of 1
Pinf$Pt
```

PATH uses the above transition probability to estimate $P$ 
for a time-step of $t=1$.

We can compare inferred transitions to the true simulated transitions, for $t=1$.

```{r}
# PATH-inferred transitions using measured branch lengths (t=1)
Pinf$P

# PATH-inferred transitions using imputed branch lengths (t=1)
Pinf.impute$P

# True transitions (t=1)
P

```



